The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-07-12 paper.tex
small fixes

\paragraph{Preservation}\label{sec:preserve}
From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite[p.~132]{HeuschkelSchmiedelLabudde2024}
Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
With these remains scheduled for reburial in 2021, digitization became essential.
For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.

\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}. 
For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.

@@

For this purpose, anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures, and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
Precisely locating the traces on the skeleton is crucial.
Accurate identification and mapping of these features allow for a detailed understanding of their implications and connections to broader anthropological inquiries.
This precision is vital for meaningful and reliable interpretations in anthropological research.
By applying the same principles used for mapping the Earth's surface \enquote{terrain}, a detailed map of the human body can be created.
This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for an objective examination, including measurements.

@@

This granular method facilitates a deep and scalable anthropological understanding.
ANNO is capable of referencing various (online) resources, even those that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.

@@

\paragraph{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
The FMA primarily focuses on depicting the entirety of human anatomy.
This broad approach results in inconsistencies in how skeletal structures are detailed, as the FMA's overarching goal to cover anatomy in general cannot sufficiently address the complexities and nuances of skeletal anatomy.
Such a wide scope also means that the anatomical knowledge pertaining to the skeleton is not depicted with the depth or precision that anthropological work demands, reflecting the limitations in existing literature on the subject.
From an anthropological perspective, the FMA ontology exhibits challenges in terms of content depth, comprehensiveness, and granularity.
Specifically, the main issues encompass a lack of cross-references between concepts, taxonomic and hierarchical inconsistencies, insufficient detail levels, and conceptual ambiguity.
These gaps result in a notable absence of detailed relationships, associations, and concepts that are crucial for a thorough anthropological understanding and analysis.
Additionally, the distinction between normal anatomy, pathology, and anatomical variants, which is essential, is not adequately addressed by the FMA's architecture.
%
Interlinks from ANNO to the FMA are added manually as it requires decions from the domain experts.
The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.

@@

\begin{itemize}
\item Specific bones can be found as a subclass of \emph{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \emph{bone tissue} (\aurl{fma}{224804}) with then varying terms such as \emph{bone tissue of hip bone} (\aurl{fma}{42854}) or \emph{bone of sacrum} (\aurl{fma}{43624}) and placed on the same level as histological structures like \emph{lamellar bone} (\aurl{fma}{224806}).
\item Different anatomical landmarks are classified taxonomically in various ways.
Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \emph{flat bone} (\aurl{fma}{7476}).
However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{anatomiedesmenschen,forensicmanual}.
To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations.
Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
\item On the other hand, another landmark, the \emph{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \emph{orifice of skull} (\aurl{fma}{53133}), which falls under a different subtree of \emph{anatomical spaces} (\aurl{fma}{5897}) in FMA.
\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \emph{anatomical point} (\aurl{fma}{9658}) that lacks coherence and detail.
\item The \emph{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
\end{itemize}

@@

%In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.%add source for that? 11 and 12 in table but no paper.
Another issue is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialized field~\citep{fma}.
For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g. on the \emph{Ossa carpi} or \emph{Ossa metatarsi}) are defined in the TA or FMA, resulting in further gaps.
A similar situation exists with respect to osteometric measurements.
Established and to some extent standardized osteometric measurement points are only found on the cranium and mandible~\citep{wesenanthropologie}.
However, since there are measurement distances on all bones, the respective start and end points must also be defined.
Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants~\citep{anatomycontribution}.
Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures~\citep{graysanatomy},
which to the greater part are not found in the FMA or the TA that primarily aim for the standardization of general (clinical) anatomy~\citep{fma}.
A work of dental anthropology even split the aforementioned Tuberculum articulare (Articular tubercle) in two structures, naming the other \enquote{articular eminence}~\citep{dentalanthropology}.
%Often exchange between these specific fields and general anatomy hardly takes place.
However, there is limited interdisciplinary coordination between the individual disciplines and general anatomy.

@@

\paragraph{Preservation}\label{sec:preserve}
From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
This is crucial because "knowledge and evidence production can only yield what the original material provides."~\cite{HeuschkelSchmiedelLabudde2024}
Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
With these remains scheduled for reburial in 2021, digitization became essential.
For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.

@@

\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
This allows researchers to "surround themselves with every shred of information about a collection or an object,"~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}. 
For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.

@@

%\section{Methodology for developing ANNO}
For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology~\citep{baumann2014, herre2010}.
The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain~\citep{baumann2014, herre2010}.
When knowledge is systematized, a set of categories is considered \emph{primitive} or \emph{basic}.
Such categories are not explicitly defined, but implicitly described by axioms~\citep{Hilbert1918}.
An example of a primitive category is \enquote{part}.
New notions can be introduced by explicit definitions based on the primitive or already defined categories~\citep{herre2010}.

@@

\paragraph{Conceptualisation}
During the conceptualisation phase~\citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
The concepts were created by generalising and classifying the domain objects.
To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
Therefore, the corresponding concepts \enquote{Bone}, \enquote{Bone part} and \enquote{Bone compound} were introduced.
Further concepts were defined to represent spatial relations and phenotypes.
Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:annodc}.

\paragraph{Axiomatization}
For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns~\citep{ODP2005, XD2016, MOMo2023}).
We instantiated specifically the GFO modules \enquote{Material objects}, \enquote{Attributives} and \enquote{Space}~\citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
For example, GFO specifies that material objects occupy space regions~\citep{Loebe2021}.
We integrated this pattern in ANNOdc by deriving the class \anno{AnatomicalStructure} from \aurl{gfo}{MaterialObject} and \anno{AnatomicalSpace} from \aurl{gfo}{SpaceRegion.}
In this way, we were able to adopt the whole axiom including the \enquote{occupies} relation between the two ANNO classes.
In a similar way, we applied the GFO pattern \enquote{point region is boundary of line region is boundary of surface region is boundary of space region}~\citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
According to the GFO relator~\citep{Loebe2018} pattern, we defined the class \anno{RelativeAnatomicalLocation} including the two roles: target (\anno{locationOf}) and reference (\anno{relativeTo}) anatomical entity. % (chapter $\ldots$).

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \enquote{Bone compound}, for example, we introduced the following axiom:
%
%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
\begin{equation*}
\anno{BoneCompound} \sqsubseteq \exists \gfo{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
\end{equation*}
%
This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
%
%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
\begin{equation*}
\anno{Phenotype} \sqsubseteq \exists \gfo{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
\end{equation*}
%
Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
%
The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
In both cases, the consistency of the ontology was proven.
Regarding use cases, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
Thereby, 14 of the core categories%
\footnote{Anatomical line, Anatomical point, Anatomical structure, Bone, Bone compound, Bone part, Measurement point, Orientation point, Phenotype, Spatial anatomical entity, Tooth, Tooth part, Phenotype and Relative anatomical location.}
and 4 of the relation types%
\footnote{hasPart, boundaryOf, locationOf and derivedFrom.}
of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
Statistical information about the ontology is shown in \cref{tab:stats}.

@@

%\paragraph{Anatomical entities}
An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
An anatomical structure refers to any anatomical division or material anatomical entity.
ANNO describes three kinds of anatomical structures: tooth structures (teeth and teeth parts), bone structures (single bones, bone parts and bone compounds), as well as the complete skeleton.
The GFO defines a material object \aurl{gfo}{MaterialObject} as a solid concrete entity that belongs to the material region of the world, has mass, consists of matter and occupies space~\citep{gfospace}.
Accordingly, we define \aurl{anno}{AnatomicalStructure} as a \aurl{gfo}{MaterialObject} in the anthropological context.
The closest equivalent in the FMA is the \emph{Material anatomical entity} \aurl{fma}{67165}, the subclass of \emph{Physical anatomical entity} \aurl{fma}{61775} that has mass.

%\paragraph{Skeleton}
The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
The TA~\citep{ta2} equates the skeleton with the \enquote{Systema skeletale (skeletal system)}, which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
ANNO does not adopt this equivalence, and \enquote{Skeleton} here refers to the entirety of a human's bones and teeth.
However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g. as specific aspects), to be integrated and referenced separately in ANNO.
The anthropologically meaningful distinction between \enquote{Skeleton} and \enquote{Systema skeletale} can be maintained, while still referencing both within the TA concept.

@@

Regarding skeletal anatomy, the representation of preservation status through the ontology would be of great use for anthropological work.
Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \enquote{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematized through ANNO, making them analyzable as well.

>>>>

\paragraph{Preservation}\label{sec:preserve}
From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
This is crucial because \enquote{knowledge and evidence production can only yield what the original material provides.}~\cite[p.~132]{HeuschkelSchmiedelLabudde2024}
Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
With these remains scheduled for reburial in 2021, digitization became essential.
By creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.

\paragraph{Access}\label{sec:access}
Anthropological skeletal material from excavations often remains undetermined due to a shortage of specialists at the respective institutions.
A digital solution would provide greater availability of human skeletal remains and allow for the efficient allocation of anthropologists and collections.
For example, the Archaeological Heritage Office of Saxony faces a significant backlog of unevaluated skeletal collections due to a shortage of anthropologists.
By digitizing these remains and using ANNO for annotation, the data can be made available to experts worldwide.
This approach allows for the efficient allocation of resources, enabling comprehensive analysis that would otherwise be infeasible.

\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
This allows researchers to \enquote{surround themselves with every shred of information about a collection or an object,}~\cite[p.~148]{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}. 
Researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.

@@

For this purpose, anthropologists use human anatomy, specifically the skeletal system and further tissues of the musculoskeletal system, to classify and describe and thus document bones, anatomical structures, and diagnostic features within the overall framework of the skeleton in a comprehensive and transparent fashion.
Precisely locating the traces on the skeleton is crucial.
Accurate identification and mapping of these features allow for a detailed understanding of their implications and connections to broader anthropological inquiries.
This precision is vital for meaningful and reliable interpretations in anthropological research.
By applying the same principles used for mapping the Earth's surface terrain, a detailed map of the human body can be created.
This map contains visible anatomical surface structures, as well as artificial objects such as measurement points or content-related or methodologically based classifications and boundaries~\citep{topo}.
ANNO represents such a map by providing an ontology for accurate and exhaustive definitions that allow to unequivocally locate these, facilitate their retrieval and furthermore serve as a basis for an objective examination, including measurements.

@@

This granular method facilitates a deep and scalable anthropological understanding.
ANNO is capable of referencing various (online) resources that are structurally and semantically diverse, such as other terminologies, ontologies, websites~\citep{atinfo} or media, aiming for synonym mediation, standardisation and conceptual clarity.

@@

\paragraph{Existing ontologies: Foundational Model of Anatomy (FMA)}\label{sec:OntologiesFMA}
Other ontologies either lack a comprehensive approach to anthropology or exhibit inconsistency in depicting knowledge within this field, since they are not designed with this specific purpose in mind.
The FMA~\citep{fma} ontology represents the physical organization of human anatomy by mapping relations to one another.
It allows the knowledge it contains to be represented in a way that is humanly comprehensible and machine-interpretable.
The FMA uses the Basic Formal Ontology (BFO) as its top-level ontology.
Corresponding to its relational nature, new higher level concepts are introduced and used for reorganizing the actual anatomical structures diverging from the TA's hierarchical structure.
Thus, in 90\,\% of the cases this leads to new creations of anatomical concepts~\citep{anatomicalterms} whereas only 1\,\% contain textual definitions~\citep{uberon}.
The FMA primarily focuses on depicting the entirety of human anatomy.
This broad approach results in inconsistencies in how skeletal structures are detailed, as the FMA's overarching goal to cover anatomy in general cannot sufficiently address the complexities and nuances of skeletal anatomy.
Such a wide scope also means that the anatomical knowledge pertaining to the skeleton is not depicted with the depth or precision that anthropological work demands, reflecting the limitations in existing literature on the subject.
From an anthropological perspective, the FMA ontology exhibits challenges in terms of content depth, comprehensiveness, and granularity.
Specifically, the main issues encompass a lack of cross-references between concepts, taxonomic and hierarchical inconsistencies, insufficient detail levels, and conceptual ambiguity.
These gaps result in a notable absence of detailed relationships, associations, and concepts that are crucial for a thorough anthropological understanding and analysis.
Additionally, the distinction between normal anatomy, pathology, and anatomical variants, which is essential, is not adequately addressed by the FMA's architecture.
%
Interlinks from ANNO to the FMA are added manually as it requires decisions from the domain experts.
The ANNO-FMA links are used in combination with unofficial FMA-TA2 links from Wikidata to automatically generate links between ANNO and the TA2.

@@

\begin{itemize}
\item Specific bones can be found as a subclass of \enquote{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \enquote{bone tissue} (\aurl{fma}{224804})
with then varying terms such as \enquote{bone tissue of hip bone} (\aurl{fma}{42854}) or \enquote{bone of sacrum} (\aurl{fma}{43624})
and placed on the same level as histological structures like \enquote{lamellar bone} (\aurl{fma}{224806}).
\item Different anatomical landmarks are classified taxonomically in various ways.
Many landmarks are described as a \enquote{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \enquote{flat bone} (\aurl{fma}{7476}).
However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{anatomiedesmenschen,forensicmanual}.
To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations.
Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
\item On the other hand, another landmark, the \enquote{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \enquote{orifice of skull} (\aurl{fma}{53133}),
which falls under a different subtree of \enquote{anatomical spaces} (\aurl{fma}{5897}) in FMA.
\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \enquote{anatomical point} (\aurl{fma}{9658})) that lacks coherence and detail.
\item The \enquote{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
\end{itemize}

@@

%In addition, the ontologies are only inadequately (TA for the last time in 2013, FMA in 2020) updated and do not undergo continuous synchronisation.%add source for that? 11 and 12 in table but no paper.
Another issue is that neither FMA nor TA are tailored to fit the requirements of anthropology or any other specialized field~\citep{fma}.
For example, articular surfaces of the individual bones are relevant anthropologically, yet not all articular surfaces (e.g. on the \latin{Ossa carpi} or \latin{Ossa metatarsi}) are defined in the TA or FMA, resulting in further gaps.
A similar situation exists with respect to osteometric measurements.
Established and to some extent standardized osteometric measurement points are only found on the cranium and mandible~\citep{wesenanthropologie}.
However, since there are measurement distances on all bones, the respective start and end points must also be defined.
Last but not least, both ontologies suffer from a lack of consideration given to anatomical variants~\citep{anatomycontribution}.
Moreover, different disciplines require a different level of detail, which is why nose surgery works with anatomical terms of very detailed anatomical structures~\citep{graysanatomy},
which to the greater part are not found in the FMA or the TA that primarily aim for the standardization of general (clinical) anatomy~\citep{fma}.
A work of dental anthropology even split the aforementioned \latin{Tuberculum articulare} (Articular tubercle) in two structures,
naming the other \emph{articular eminence}~\citep{dentalanthropology}.
%Often exchange between these specific fields and general anatomy hardly takes place.
However, there is limited interdisciplinary coordination between the individual disciplines and general anatomy.

@@

\paragraph{Preservation}\label{sec:preserve}
From the moment of excavation, skeletal remains are subject to wear and tear, diminishing their informational value.
Transferring the examination to digital models reduces physical handling, minimizes deterioration, and ensures that data is preserved and accessible for future research, even in cases of repatriation or reburial.
This is crucial because \enquote{knowledge and evidence production can only yield what the original material provides.}~\cite{HeuschkelSchmiedelLabudde2024}
Exemplifying ANNO's underlying utility in this regard is the Rödelheim skeletal collection curated and researched by the Department of Historical Anthropology and Human Ecology at the University of Göttingen, consisting of over 200 skeletal remains from Napoleon's army.
With these remains scheduled for reburial in 2021, digitization became essential.
For instance, by creating digital 3D models via AnthroWorks3D, and annotating them based on the ANNO ontological framework, the models were converted into research data that is analyzable across studies, despite the physical remains being reburied.

@@

\paragraph{Comparability and Transparent Documentation and Analysis}\label{sec:comp+transp} 
Existing data recording systems in anthropology are often individualized, leading to compatibility issues that hinder comparative analysis and comprehensive intra- and interdisciplinary research~\cite{HeuschkelSchmiedelLabudde2024}.
A digital solution combining ANNO, a standardized ontology for skeletal data, and AnthroWorks 3D, a 3D editor for creating and annotating digital bone models, provides unprocessed and undistorted information, making it comprehensible, reproducible, and sustainably documented.
It also supports flexible and objective documentation of bone parts or fragments~\cite{aw3dcidoc}.
This allows researchers to \enquote{surround themselves with every shred of information about a collection or an object,}~\cite{Palkovich.2001} with future possibilities such as integration into collection management systems and the application of AI analysis~\cite{HeuschkelSchmiedelLabudde2024}. 
For instance, researchers comparing skeletal remains from different regions can use ANNO to standardize their annotations, ensuring that data from disparate sources is compatible.
For instance, if two teams are studying bone wear patterns to infer lifestyle differences using different skeletal collections, ANNO allows them to document their findings in a uniform manner, facilitating direct comparisons and more robust conclusions.

@@

%\section{Methodology for developing ANNO}
For the development of ANNO, we applied the onto-axiomatic method, a combination of the axiomatic method with a top-level ontology~\citep{baumann2014, herre2010}.
The axiomatic method comprises principles for developing theories or formal knowledge bases, which aim at the foundation, systematisation and formalisation of a knowledge domain~\citep{baumann2014, herre2010}.
When knowledge is systematized, a set of categories is considered \emph{primitive} or \emph{basic}.
Such categories are not explicitly defined, but implicitly described by axioms~\citep{Hilbert1918}.
An example of a primitive category is \emph{part}.
New notions can be introduced by explicit definitions based on the primitive or already defined categories~\citep{herre2010}.

@@

\paragraph{Conceptualisation}
During the conceptualisation phase~\citep{herre2010}, the core concepts (categories, classes) and relations were introduced that form ANNOdc (domain-core ontology) (\cref{fig:annodc}).
The concepts were created by generalising and classifying the domain objects.
To answer the competence questions of the first use case, for example, a distinction must be made between whole bones, bone parts and composite bone structures.
Therefore, the corresponding concepts \emph{Bone}, \emph{Bone part} and \emph{Bone compound} were introduced.
Further concepts were defined to represent spatial relations and phenotypes.
Furthermore, we identified relations (e.g., partOf, boundaryOf or locationOf) relevant to capture axioms covering the defined use cases, see \cref{fig:annodc}.

\paragraph{Axiomatization}
For the axiomatization and formal foundation of ANNO we utilised GFO as a top-level ontology and reused its categories, relations, axioms and modules (as a kind of Ontology Design Patterns~\citep{ODP2005, XD2016, MOMo2023}).
We instantiated specifically the GFO modules \emph{Material objects}, \emph{Attributives} and \emph{Space}~\citep{Burek2020, Loebe2021, Loebe2018} and adapted them according to the requirements of the use cases.
Domain concepts introduced in the conceptualisation phase were embedded in GFO, i.e., defined as subcategories (subclasses) of certain GFO categories (fig. 1).
The required relations between the concepts were then either adopted from GFO or derived (specialised) from the corresponding GFO relation.
For example, GFO specifies that material objects occupy space regions~\citep{Loebe2021}.
We integrated this pattern in ANNOdc by deriving the class \anno{AnatomicalStructure} from \aurl{gfo}{MaterialObject} and \anno{AnatomicalSpace} from \aurl{gfo}{SpaceRegion.}
In this way, we were able to adopt the whole axiom including the \emph{occupies} relation between the two ANNO classes.
In a similar way, we applied the GFO pattern \emph{point region is boundary of line region is boundary of surface region is boundary of space region}~\citep{baumann2016} to the ANNO categories anatomical point, line, surface and space.
According to the GFO relator~\citep{Loebe2018} pattern, we defined the class \anno{RelativeAnatomicalLocation} including the two roles: target (\anno{locationOf}) and reference (\anno{relativeTo}) anatomical entity. % (chapter $\ldots$).

Further axioms were introduced to precisely define certain categories.
In addition to an explicit textual definition of the category \emph{Bone compound}, for example, we introduced the following axiom:
%
%\anno{BoneCompound}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{hasPart} some (\anno{Bone} or \anno{BoneCompound} or \anno{BonePart}))
\begin{equation*}
\anno{BoneCompound} \sqsubseteq \exists \gfo{hasPart}.(\anno{Bone} \sqcup \anno{BoneCompound} \sqcup \anno{BonePart})
\end{equation*}
%
This axiom states that a bone compound consists of further bone compounds, individual bones or bone parts.
Another example axiom postulates that phenotypes can be derived from anatomical entities or anatomical properties:
%
%\anno{Phenotype}: \aurl{rdfs}{subClassOf} (\aurl{gfo}{derivedFrom} some (\anno{AnatomicalEntity} or \anno{AnatomicalProperty}))
\begin{equation*}
\anno{Phenotype} \sqsubseteq \exists \gfo{derivedFrom}.(\anno{AnatomicalEntity} \sqcup \anno{AnatomicalProperty})
\end{equation*}
%
Finally, a spreadsheet template was developed for domain experts to specify domain-specific entities (ANNOds), strictly compliant with ANNOdc, see \cref{sec:annodc}.
This means that the domain experts had to insert more specific categories (single categories or category trees) under defined ANNOdc core categories and to describe them by suitable annotations and relations introduced in ANNOdc.
%
The resulting OWL 2 ontology was generated using SMOG~\citep{smog} and validated by the HermiT~\citep{hermit} and Pellet~\citep{pellet} reasoners and SHACL shapes, see \cref{sec:annods}.
In both cases, the consistency of the ontology was proven.
Regarding use cases, we integrated the ontology into the \aw{} software~\citep{aw3d} and were able to successfully realise all three intended use cases.
Thereby, 14 of the core categories%
\footnote{Anatomical line, Anatomical point, Anatomical structure, Bone, Bone compound, Bone part, Measurement point, Orientation point, Phenotype, Spatial anatomical entity, Tooth, Tooth part, Phenotype and Relative anatomical location.}
and 4 of the relation types%
\footnote{hasPart, boundaryOf, locationOf and derivedFrom.}
of ANNOdc were utilised and their subcategories and specific relations were dynamically queried (according to the three-ontology method).
Statistical information about the ontology is shown in \cref{tab:stats}.

@@

%\paragraph{Anatomical entities}
An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
An anatomical structure refers to any anatomical division or material anatomical entity.
ANNO describes three kinds of anatomical structures: tooth structures (teeth and teeth parts), bone structures (single bones, bone parts and bone compounds), as well as the complete skeleton.
The GFO defines a material object \aurl{gfo}{MaterialObject} as a solid concrete entity that belongs to the material region of the world, has mass, consists of matter and occupies space~\citep{gfospace}.
Accordingly, we define \aurl{anno}{AnatomicalStructure} as a \aurl{gfo}{MaterialObject} in the anthropological context.
The closest equivalent in the FMA is the \enquote{Material anatomical entity} \aurl{fma}{67165}, the subclass of \enquote{Physical anatomical entity} \aurl{fma}{61775} that has mass.

%\paragraph{Skeleton}
The largest or most comprehensive skeletal anatomical structure, the human skeleton, is the framework composed of all the bones and teeth of a human being.
The TA~\citep{ta2} equates the skeleton with the \latin{Systema skeletale} (skeletal system), which refers to the passive part of the musculoskeletal system, including the articulated bony skeleton with teeth, as well as cartilage, ligaments, and other joints.
ANNO does not adopt this equivalence, and \emph{Skeleton} here refers to the entirety of a human's bones and teeth.
However, since teeth constitute their own tissue, they are treated separately, allowing all other components of the skeletal system, such as cartilage or ligaments (as tissues) or joints (e.g. as specific aspects), to be integrated and referenced separately in ANNO.
The anthropologically meaningful distinction between \emph{Skeleton} and \latin{Systema skeletale} can be maintained, while still referencing both within the TA concept.

@@

Regarding skeletal anatomy, the representation of preservation status through the ontology would be of great use for anthropological work.
Furthermore, anatomical terms describing aspects of bone morphology or function, such as the Latin word \latin{processus} which appear numerous times as part of anatomical designations indicating a projection, could be reviewed and flexibly systematized through ANNO, making them analyzable as well.
